Geometric variational problems

### 2020-11-12 NSF2020/Description/specificaim1_ver2.tex

Here we start with a particle-based bilayer in a specific non-equilibrium shape that consists of only one component of the displacement in figure~\ref{fig:deformations}, 
and then evolve the particle system
%bilayer by steepest gradient descent 
with respect to the HARP functional:
\begin{equation}
\label{SGD}   
\frac{\dif {\bf a}_i}{\dif t} = {\bf F}_i,    \quad  \frac{\dif {\bf R}_i}{\dif t}  = {\bf R}_i A({\bf w}_i),\quad i = 1, \dots, N.
\end{equation}
where that particle centers ${\bf a}_i$ and rotation matrices ${\bf R}_i$ determine the particle configuraton; 
$A({\bf w}_i)$ is the skew-symmetric matrix with axial vector ${\bf w}_i$.
%We solve \eqref{SGD} numerically  using an Adams–Bashforth linear multistep method. 
The procedure involves solving \eqref{SL} based off of the current particle configuration.
We then use the solution $u$ to calculate the collection of forces and torques \eqref{forceandtorque}, and update the particle configuration
through \eqref{SGD}.

\begin{wrapfigure}[12]{l}{0.32\textwidth}
\centerline{\includegraphics[width=0.32\textwidth]{Figures/SaddleSplayDiagram.pdf}}
\caption{\label{fig:saddle_splay} \footnotesize Determination of
  saddle splay modulus.}
\end{wrapfigure}
%
The evolution \eqref{SGD} satisfies the energy dissipation law $\frac{\dif }{\dif t} \Phi \leq 0$
and this stabilizes the evolving bilayer shape.
Therefore, we reconstruct an evolving monolayer dividing surface $\Sigma$ and director field ${\bf n}$ by
interpolating the particle centers and orientations. Using \eqref{ansatz3}, we calculate a continuum energy $E$ from the interpolated shapes.

@@

We propose to conduct calculations similar to the example of figure~\ref{fig:flattening} for other elastic moduli such as the effective saddle splay modulus $\KG$:
%for a wide range of particle parameters and physical parameters in HARP to 
%
%We will also determine the effective saddle splay modulus $\KG$. 
The gradient descent technique is ineffective for measuring $\KG$ because
the saddle splay energy is largely invariant under shape changes.
In the simplified Canham-Helfrich formulation, saddle splay energy is an indicator of topological transitions, thanks to the  Gauss-Bonnet theorem
\cite{TerziDeserno17}.
PI-RR showed that saddle splay acts as a topological indicator even in the presence of nonzero tilt \cite{RyKlYaCo16}.

To evaluate $\KG$, we will combine the string method from PI-Ryham's work on fusion \cite{RyKlYaCo16} with
the present particle simulations. 
%The string method is a numerical scheme that finds
%least energy pathways separating energy basins \cite{doi:10.1063/1.2720838}.
%Using the string method, we will evaluate the transition energies between the intact pancake shape,
%and one with a pore. Pore formation in a single bilayer decreases the integral of saddle splay from $4\pi$
%to $0$. We can, in principle, detect this drop in energy because we have already accounted for the other deformations. 
%
%%We will also analyze $\KG$ using a helical geometry. In a helicoid with inclination $m$, the director
%%gradient is $\nabla n = \begin{pmatrix} 0 & m(m^2+x_1^2)^{-1} \\ m(m^2+x_1^2)^{-1} & 0\end{pmatrix}$ where $x_1$ is the distance to the central
%%axis. We observe that saddle splay, twist and tilt are zeros, and $\det \nabla n = m^2(m^2+x_1^2)^{-2}$.
%%In the simulations, the particles forming the helicoid will be confined to two, concentric cylinders. The
%%continuum energy $\tfrac{1}{2}\KG h(m x_1(m^2+x_1^2)^{-1} + \tan^{-1}(x_1/m)) |_{r_1}^{r_2}$ where $h$, $r_1$ and $r_2$
%%are the cylinder heights, inner cylinder radius, and outer cylinder radius, respectively. 
%
The twist deformation is a fully three-dimensional deformation. 
(Specific Aim 3 address outstanding implementation issues like three-dimensional boundary integral solvers.) 
It is closely related to the assumption of lateral fluidity in membranes \cite{Hamm2000}. Note that tilt is locally non-zero whenever twist is non-zero.
That is because the surface gradient $\nabla {\bf n}$ equals the second fundamental form, and is therefore symmetric, whenever ${\bf n} = {\bf N}$ locally.

@@

\subsubsection{HK elastic energy form}
It may come as a surprise then that the field of membrane continuum mechanics still lacks consensus as to
 whether \eqref{ansatz3} contains a complete list of consistent deformations.
\textbf{(i)} Researcher have assumed that $\KT = 0$ \cite{Hamm2000, TerziDeserno17, C9SM02079A, PhysRevE.102.042406}.
  Based on \eqref{eq:twist_energy}, a value $\KT = 0$ leads to the
  unphysical conclusion that there are bilayers with $\nabla {\bf n} \to \infty$ but $E \to 0.$ 
  \textbf{(ii)}
  Recently, \cite{TerziDeserno17} derived a tilt curvature term that was neglected from the HK analysis \cite{Hamm2000}.
  Later, \cite{C9SM02079A} 
  and \cite{PhysRevE.102.042406} independently identified an inconsistency \cite{TerziDeserno17} arising
  from a transversal tilt invariance assumption.
  The PI Ryham showed,  that the tilt vector $T = {\bf n}/{\bf n}\cdot {\bf N} - {\bf N}$ leads to unphysical cusps when constraining membrane thickness  \cite{RyKlYaCo16}.  
  \textbf{(iii)}
  Theoretical analysis of lipid phase transitions predict a negative saddle-splay modulus around $-8$ \kBT\;
\cite{SIEGEL2004366,SIEGEL20085200}. Researchers point out, however, that this theoretical value requires a larger energy 
barrier for monolayer fusion than is found by experiment \cite{FrRoPi17,Tran7106,TerziDeserno17}.
Furthermore, the stability requirements of \eqref{ansatz3} are $\KB > 0$, $K_T > 0$, $\KG \geq -4\KB$
and $\KG \geq -4\KT$. Molecular dynamics investigations find a twist modulus about 1 \kBT \cite{LeVeWa14},
suggesting a somewhat smaller magnitude for $\KG$.

% The form of the elastic energy density \eqref{ansatz3} is the same as
%the Oseen-Frank energy density for nematic liquid crystals \cite{ANDRIENKO2018520,Tran7106,Helfrich73}.   In fact,  
%a lipid monolayer acts as one layer in a smectic  phase \cite{REYESMATEO1995978,Rangamani20140463,PhysRevLett.113.248102}. 
%The constants $\KB$, $\KT$  and $\KG$ play the same role as the Frank constants $K_1$, $K_2$ and $K_{24}$
%liquid crystal theory. However, the ``bend'' distortion from nematics  
%is not present in monolayers since the directors have no dependence in the normal direction.
%There is also no ``spontaneous twist'' in monolayers due to invariance under mirror reflection.

  We suspect the reason why, after twenty years, the field has yet to reach a consensus 
  as to the form of \eqref{ansatz3} has to do with the Taylor expansions used to derive the elastic energy density. 
  Past researchers expressed the strain tensor in
  terms of the orthonormal frame $\{{\bf e}_1, {\bf e}_2, {\bf N}\}$ with surface tangent vectors ${\bf e}_i$ and surface normal ${\bf N}$.  
  In this frame, there is a coupling between the gradients of all three terms ${\bf x}_0,$ $\zeta$ and ${\bf n}$ of the surface deformation field.
  %\eqref{LMdeformation}.
  Subtle differential geometric arguments are needed to evaluate the terms in the Taylor expansion,
  and these couplings can lead to an energy density with more than ten parameters, for example \cite{PhysRevE.102.042406}.

  We will revisit the HK analysis to see if any of the aforementioned inconsistencies can be resolved.
  As a main analytical innovation, we have found that expanding the strain tensor in
  terms of the orthonormal frame $\{{\bf e}'_1, {\bf e}'_2, n\}$ with ${\bf e}'_i$ perpendicular to ${\bf n}$
  greatly simplifies the expansion because it decouples the gradient terms. As an illustration, the prominent works
  \cite{TerziDeserno17, PhysRevE.102.042406, Hamm2000, C9SM02079A} utilize an
  incompressibility condition for the lipid elongation function $\zeta$ from \eqref{LMdeformation} that is only approximately true. 
  However, we prove that when expressed in the $\{{\bf e}'_1, {\bf e}'_2, n\}$ frame, the incompressility condition is \emph{exact}
  and is given by a Steiner-type polynomial in terms of $\Div {\bf n}$ and $\det \nabla {\bf n}$ \cite{Fe59}.
  Aside from the special case of zero tilt, \cite{Hamm2000} and coworkers seem to have been unaware of this result.

>>>>

Here we start with a particle-based bilayer in a specific non-equilibrium shape that involves only one of the components of the displacement in Figure~\ref{fig:deformations}, and then evolve the particle system according to the time integration for \eqref{eq:stokes}.
Actually,  elastic properties are independent of dissipation, and so we can forgo solving for fluid velocity and by setting the translational
velocity and angular velocity directly proportional to force and torque. 
%bilayer by steepest gradient descent 
%with respect to the HARP functional:
%\begin{equation}
%\label{SGD}   
%\frac{\dif {\bf a}_i}{\dif t} = {\bf F}_i,    \quad  \frac{\dif {\bf R}_i}{\dif t}  = {\bf R}_i A({\bf w}_i),\quad i = 1, \dots, N.
%\end{equation}
%where that particle centers ${\bf a}_i$ and rotation matrices ${\bf R}_i$ determine the particle configuraton; 
%$A({\bf w}_i)$ is the skew-symmetric matrix with axial vector ${\bf w}_i$.
%We solve \eqref{SGD} numerically  using an Adams–Bashforth linear multistep method. 
%The procedure involves solving \eqref{SL} based off of the current particle configuration.
%We then use the solution $u$ to calculate the collection of forces and torques \eqref{forceandtorque}, and update the particle configuration
%through \eqref{SGD}.

\begin{wrapfigure}[12]{l}{0.32\textwidth}
\centerline{\includegraphics[width=0.32\textwidth]{Figures/SaddleSplayDiagram.pdf}}
\caption{\label{fig:saddle_splay} \footnotesize Determination of
  saddle splay modulus.}
\end{wrapfigure}
%
In either case there is a dissipation of the total potential \eqref{eq:total_poten}, and this stabilizes the evolving bilayer shape.
Therefore, we reconstruct an evolving monolayer dividing surface $\Sigma$ and director field ${\bf n}$ by
interpolating the particle centers and orientations. Using \eqref{ansatz3}, we calculate a continuum energy $\mathbcal{W}$ from the interpolated shapes.

@@

We propose to conduct calculations similar to the example of figure~\ref{fig:flattening} for other elastic moduli such as the effective twist modulus $\KT$.
Molecular dynamics investigations find a twist modulus about 1 \kBT \cite{LeVeWa14}, and here the specific non-equilibrium shape consists of a single layer of amphiphilic particles on a hydrophobic substrate as illustrated by Figure~\ref{fig:deformations}D. Having a nonzero twist requires nonzero tilt, because the surface gradient of the lipid directoy equals the second fundamental form, and is therefore symmetric, whenever tilt is zero locally. The twist deformation is a fully three-dimensional deformation. \S subsec:specific_aim_3 addresses outstanding implementation issues like three-dimensional boundary integral equation solvers.

%for a wide range of particle parameters and physical parameters in HARP to 
%
%We will also determine the effective saddle splay modulus $\KG$. 
The gradient descent technique is ineffective for measuring the saddle splay modulus $\KG$ because
the saddle splay energy is largely invariant under shape changes.
To evaluate $\KG$, we will combine the present particle simulations with the string method 
from PI-RR's work on membrane fusion \cite{RyKlYaCo16}. The string method is a numerical scheme that finds
least energy pathways separating energy basins \cite{doi:10.1063/1.2720838}. 
In the simplified Canham-Helfrich formulation, saddle splay energy is an exact indicator of topological transitions, 
thanks to the  Gauss-Bonnet theorem \cite{TerziDeserno17}.
More generally, PI-RR has shown that saddle splay acts as a topological indicator even in the presence of nonzero tilt \cite{RyKlYaCo16}. 
As a result, saddle splay can quantified from the transition energies of a least energy path of pore formation (Figure \ref{fig:saddle_splay}).   
%Using the string method, we will evaluate the transition energies between the intact pancake shape,
%and one with a pore. Pore formation in a single bilayer decreases the integral of saddle splay from $4\pi$
%to $0$. We can, in principle, detect this drop in energy because we have already accounted for the other deformations. 
%
%%We will also analyze $\KG$ using a helical geometry. In a helicoid with inclination $m$, the director
%%gradient is $\nabla n = \begin{pmatrix} 0 & m(m^2+x_1^2)^{-1} \\ m(m^2+x_1^2)^{-1} & 0\end{pmatrix}$ where $x_1$ is the distance to the central
%%axis. We observe that saddle splay, twist and tilt are zeros, and $\det \nabla n = m^2(m^2+x_1^2)^{-2}$.
%%In the simulations, the particles forming the helicoid will be confined to two, concentric cylinders. The
%%continuum energy $\tfrac{1}{2}\KG h(m x_1(m^2+x_1^2)^{-1} + \tan^{-1}(x_1/m)) |_{r_1}^{r_2}$ where $h$, $r_1$ and $r_2$
%%are the cylinder heights, inner cylinder radius, and outer cylinder radius, respectively. 
%

@@

\subsubsection{HK elastic energy form}
The field of membrane continuum mechanics still lacks consensus as to whether \eqref{ansatz3} is the appropriate functional for bilayer energy.
\textbf{(i)} Researcher have assumed that $\KT = 0$ because of the relationship between twist the assumption of lateral fluidity in membranes 
\cite{Hamm2000, TerziDeserno17, C9SM02079A, PhysRevE.102.042406}. 
A value $\KT = 0$, however, makes \eqref{ansatz3} noncoercive and leads to the nonphysical conclusion that there are bilayers 
with $\nabla {\bf n} \to \infty$ but $\mathcal{W} \to 0.$ 
  \textbf{(ii)}
Recently, \cite{TerziDeserno17} derived a tilt curvature term that was neglected from the HK analysis \cite{Hamm2000}.
Later, \cite{C9SM02079A} 
and \cite{PhysRevE.102.042406} independently identified an inconsistency an the argument used by \cite{TerziDeserno17} arising
from a transversal tilt invariance assumption.
In \cite{RyKlYaCo16}, the PI RR and collaborators showed that the tilt vector leads to unphysical cusps depending on how one accounts for membrane thickness.  
  \textbf{(iii)}
  Theoretical analysis of lipid phase transitions predict a negative saddle-splay modulus around $-8$ \kBT\;
\cite{SIEGEL2004366,SIEGEL20085200}. Researchers point out, however, that this theoretical value requires a larger energy 
barrier for monolayer fusion than is found by experiment \cite{FrRoPi17,Tran7106,TerziDeserno17}.
Furthermore, the stability conditions of \eqref{ansatz3} require $\KG \geq -4\min\{\KB,\KT\}$, suggesting a smaller magnitude for $\KG$.

% The constants $\KB$, $\KT$  and $\KG$ play the same role as the Frank constants $K_1$, $K_2$ and $K_{24}$
% liquid crystal theory. However, the ``bend'' distortion from nematics  
% is not present in monolayers since the directors have no dependence in the normal direction.
% There is also no ``spontaneous twist'' in monolayers due to invariance under mirror reflection.

  We will revisit the HK analysis to see if any of the aforementioned inconsistencies can be resolved.
  Past researchers expressed the strain tensor in terms of monolayer tangent plane.   
  In this frame, however, there is a coupling between the gradients of the monolayer surface, lipid direction, and chain length,
  and accounting for these couplings by Taylor expansion  
  can lead to an energy density with an unruly number of parameters \cite{PhysRevE.102.042406}.

The form of the elastic energy density \eqref{ansatz3} is the same as
  the Oseen-Frank energy density for nematic liquid crystals \cite{ANDRIENKO2018520,Tran7106,Helfrich73}.   In fact,  
  a lipid monolayer acts as one layer in a smectic  phase \cite{REYESMATEO1995978,Rangamani20140463,PhysRevLett.113.248102}. 
  Motivated by liquid crystal theory, we have found that the gradient terms completely decouples when 
  expanding the strain tensor, not in terms of the monolayer tangent plane, but in terms of a plane perpendicular to ${\bf n}$. 
  Using this observation, we prove an exact identity 
  that gives the incompressility condition by a Steiner-type polynomial in $\Div {\bf n}$ and $\det \nabla {\bf n}$ \cite{Fe59}.
  In contrast, the works \cite{TerziDeserno17, PhysRevE.102.042406, Hamm2000, C9SM02079A} utilize an
  approximate identity for incompressibility, and so we can offer a considerably more accurate analysis of monolayer energy.
